Master Thesis

### 2013-05-13 ig.tex
Wrote about negation in Frigram

The nodes of the syntactic tree are also decorated with
features. Interaction grammars are a formalism enabling us to define
sets of the structures described above, the \emph{syntactic
  trees}. Similar to Tree Adjoining Grammars, interaction grammars are
formulated in terms of a set of elementary structures which can combine
to produce the final output structures. However, unlike in tree
adjoining grammars, the output structure is not constructed from the
elementary structures via some set of algebraic operations (substitution
and adjunction in case of TAG). Instead, \emph{polarized tree
  descriptions} (PTDs), the elementary structures of interaction
grammars, impose constraints on the final structure and a structure is
said to be generated by some PTDs if it is a minimal structure
satisfying those constraints (it is a \emph{minimal model}).

@@

The formalism also allows us to specify \emph{immediate precedence} and
general \emph{dominance}. The latter is useful for modelling unbounded
dependencies such as those between a relative pronoun and its trace in
some embedded clause of the relative clause, but it is not used in our
present example. Finally, one kind of constraint that is used in our
example is the orange rectangle in node (3,0), which requires that the
interpretation of (3,0) is the right-most daughter amongst its siblings.

Now that we have covered the structural constraints imposed by the tree
descriptions, we will turn our attention to the defining characteristic
of interaction grammars, the polarities. As you have noticed on
Figure~\ref{fig:ig-eptds}, some of the features in our PTDs are
annotated with special symbols and colors. These denote the different
polarities and are used by the formalism for two distinct purposes: the
positive ($\textcolor{red}{\rightarrow}$) and negative
($\textcolor{blue}{\leftarrow}$) polarities are used to model the
resource sensitivity of language, while the virtual polarities
($\textcolor{purple}{\sim}$) are used to match against the
surrounding context.

The way the polarities are handled is that every model (output syntactic
tree) is required to have only saturated polarities on its features
(i.e. no positive, negative or virtual polarities). Whenever more than
one node of the PTD is mapped onto a single node of the syntactic tree,
the polarities of each feature are combined. The combination mechanism
allows to combine a positively polarized feature with a negatively
polarized one to yield a saturated one. Virtual features can be factored
out only by combining with a saturated polarity (they are somewhat
analogous to the $=_{c}$ constraints of LFG \cite{kaplan1982lexical}).

@@

>>>>

The nodes of the syntactic tree are also decorated with
features.

Interaction grammars are a formalism enabling us to define sets of the
structures described above, the \emph{syntactic trees}. Similar to Tree
Adjoining Grammars, interaction grammars are formulated in terms of a
set of elementary structures which can combine to produce the final
output structures. However, unlike in tree adjoining grammars, the
output structure is not constructed from the elementary structures via
some set of algebraic operations (substitution and adjunction in case of
TAG). Instead, \emph{polarized tree descriptions} (PTDs), the elementary
structures of interaction grammars, impose constraints on the final
structure and a structure is said to be generated by some PTDs if it is
a minimal structure satisfying those constraints (it is a \emph{minimal
  model}).

@@

The formalism also allows us to specify \emph{immediate precedence} and
(large) \emph{dominance}. The latter is useful for modelling unbounded
dependencies such as those between a relative pronoun and its trace in
some embedded clause of the relative clause, but it is not used in our
present example.

Finally, one kind of constraint that is used in our example is the
orange rectangle in node (3,0), which requires that the interpretation
of (3,0) is the rightmost daughter amongst its siblings.

Now that we have covered the structural constraints imposed by the tree
descriptions, we will turn our attention to the defining characteristic
of interaction grammars, the polarities. As you have noticed on
Figure~\ref{fig:ig-eptds}, some of the features in our PTDs are
annotated with special symbols and colors. These denote the different
polarities and are used by the formalism for two distinct purposes: the
positive ($\textcolor{red}{\rightarrow}$) and negative
($\textcolor{blue}{\leftarrow}$) polarities are used to model the
resource sensitivity of languages, while the virtual polarities
($\textcolor{purple}{\sim}$) are used to for pattern matching against
the context.

The way the polarities are handled is that every model (output syntactic
tree) is required to have only saturated polarities on its features
(i.e. no positive, negative or virtual polarities). Whenever more than
one node of the PTD is mapped onto a single node of the syntactic tree,
the polarities of each feature are combined. The combination mechanism
allows us to combine a positively polarized feature with a negatively
polarized one to yield a saturated one. Virtual polarities can only be
eliminated by combining them with a saturated polarity (they are
analogous to the $=_{c}$ constraints of LFG \cite{kaplan1982lexical}).

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.25]{ig-neg.pdf}
  \caption{\label{fig:ig-neg} EPTDs of $aucun$ and $ne$ demonstrating
    the way polarities are used to model French negation.}
\end{figure}

We will briefly discuss how interaction grammars, and Frigram in
particular, handle a tricky linguistic phenomenon in an interesting way.
In French, negation is signalled by the particle $ne$ which must be
accompanied by one of several designated determiners, pronouns or
adverbs. The particle $ne$ assumes a position right before the inflected
verb, but its partner, e.g. the determiner $aucun$ can appear in a rich
variety of positions:

\begin{exe}
  \ex \label{ex:aucun-suj} Aucun tatou ne court.
  \ex \label{ex:aucun-obj} Jean n'aime l'odeur d'aucun tatou.
  \ex \label{ex:aucun-bad} * Le tatou qu'aucun loup chasse ne court.
  \ex \label{ex:aucun-good} Le tatou qu'aucun loup ne chasse court.
\end{exe}

Sentences (\ref{ex:aucun-suj}) and (\ref{ex:aucun-obj}) demonstrate that
the $aucun$ determiner can be used both in the subject and object
positions, as a determiner of either one of the arguments directly or of
a noun phrase which complements one of them. However, it is not
admissible to place the determiner in an embedded clause as it is in
(\ref{ex:aucun-bad}). If the determiner $aucun$ is used in an embedded
clause, it is the embedded clause itself that must be negated, as in
sentence (\ref{ex:aucun-good}).

This example demonstrates two points of interest to us.

First, it uses a new polarized feature, $neg$, to model the fact that
the two wordforms which enable negation must always co-occur. The
particle $ne$ provides a positive $neg$ feature to the clause in which
it occurs and this positive polarity must be saturated by a negative
polarity contributed by one of the possible partner words, such as
$aucun$. This shows that the usefulness of positive and negative
polarities extends beyond the $cat$/$funct$ pair that we have already
seen and which is used to model predicate/argument structure in a
similar way as was already done in tree adjoining grammars or categorial
grammars.

Second, the determiner $aucun$ can occur deep, e.g. inside a complement
of one of the arguments, but it cannot occur everywhere, as is
demonstrated by (\ref{ex:aucun-bad}), where we try to put the determiner
inside another embedded clause. There is a way to express this kind of
constraint in interaction grammars that we have not revealed
yet. Whenever we use a (large) dominance (a top-down green dashed line),
we usually want the range of nodes that this dominance relation can
cross to be somehow restricted. In interaction grammars, we can
associate a feature structure with a dominance constraint which will
force all the nodes that will span the distance between the two nodes
linked by the dominance constraint to unify with that feature
structure.

In the notation used in our illustrations (which were generated by the
Leopar parser\footnote{\url{http://leopar.loria.fr/}}), this is conveyed
by putting all the restricting features as virtual features in a new
node and splicing the new node in between the two nodes linked by the
dominance constraint before (see the node (0,4) in
Figure~\ref{fig:ig-neg}, which merely serves to ensure that all the
nodes on the path from (0,3) to (0,2) have either $np$ or $pp$ as the
value of their $cat$ feature). When using this notation, the (large)
dominance constraint should then be read as not only stating that the
top node must be an ancestor to the bottom node, but also that any
intermediate nodes between the ancestor and successor must match the
ancestor's features.
